---
title: "Exploring hrbrthemes"
output: html_notebook
---

# Introduction
These themes emphasise careful font design, and follow a nice, generally minimalist design aesthetic otherwise. Here I deom the different themes in the hrbrthemes package, for later review and discussion.


# Load Libraries
```{r}
require(tidyverse)
require(ggplot2)

# For testing
library(extrafont)
library(hrbrthemes)
```

# Load Data
```{r}
sibfldnm <- 'Derived Data'
parent <- dirname(getwd())
sibling <- file.path(parent,sibfldnm)
fn = 'Shellfish data 2015 2018.csv'
path <- file.path(sibling, fn)
coli_data <- read_csv(path)

coli_data <- coli_data %>%
  mutate_at(4:8, factor) %>%
  mutate(Class = factor(Class,levels = c( 'A', 'CA', 'CR',
                                         'R', 'P', 'X' ))) %>%
  mutate(Tide = factor(Tide, levels = c("L", "LF", "F", "HF",
                                        "H", "HE", "E", "LE")))
coli_limits <- list(low = 0, gmlow=14, p90low=31, gmhigh=88, p90high=163, high= 5000)

freq_data<- coli_data %>%
  filter(! is.na(ColiVal)) %>%
  mutate(lvls4 = cut(ColiVal, coli_limits, labels = c('low', 'gmlow', 'p90low', 'gmhigh', 'p90high')))
freq_data <- freq_data %>% 
  select(-ColiScore, -RawColi, -LCFlag, -RCFlag, -ColiVal)
rm(coli_data, coli_limits, fn, parent, sibling, sibfldnm, path)
```


# Import hrbrthemes Fonts
I probably don't need to do this every time, but this is for completeness.  Despite the warnings, have already installed all these fonts in the  windows font folder
```{r}
import_econ_sans()
import_plex_sans()
import_public_sans()
import_roboto_condensed()
import_tinyhand()
import_titillium_web()
```

#Testing another set of graphic themes and fonts.
```{r}
mo_agg_dat <- freq_data %>%
  mutate(MONTH = factor(MONTH, labels = c('jan','feb','mar','apr','may','jun','jul','aug','sep','oct','nov','dec'))) %>%
  count(MONTH, lvls4)
 
plt <- ggplot(mo_agg_dat, aes(MONTH, n, fill=lvls4)) +
   geom_bar(stat = "identity", position='fill')
   
plt + theme_ipsum_rc()
   
```
## Other versions
hrbrthemes comes with multiple default themes focused on typography, including
Abreviation |  Example Function Call  |  Font Family
------------|-------------------------|---------------------
"rc"        |  theme_ipsum_rc()       |Roboto Condensed
""          |  theme_ipsum()          |Arial Narrow
"es"        |  theme_ipsum_es()       |Econ Sans Condensed
"ps"        |  theme_ipsum_ps()       |BM Plex Sans
"pub"       |  theme_ipsum_pub()      |Public Sans
"tw"        |  theme_ipsum_tw()       |Titillium Web
"tinyhand"  |  theme_tinyhand()       | Tinyhand
 
### Reversed Shading
"Modern R Studio" Theme
```{r}
plt + theme_modern_rc()  #also try theme_ft_rc()
```
 
```{r}
plt + theme_ipsum_es()
```
```{r}
plt + theme_ipsum_pub()
```
  
```{r}
plt  + theme_ipsum_tw()
```
And, an especially odd-looking example , useable for draft graphics in powerpoint, perhaps.  It reminds me  a bit of the look using the xkcd fonts, but it's harder to read, despite better kerning.
```{r}
plt  + theme_tinyhand(base_family = "BF Tiny Hand") +
  theme(axis.text.x = element_text(angle=90, hjust = .25, vjust = .25))
```

